---
title: "Final Project Proposal"
author: "Mohit Khanna, Eric A. Scuccimarra, Shikhar Khanna"
date: "13/07/2019"
output:
  html_document: 
    toc: yes
  pdf_document: default
urlcolor: cyan
---

# Project Team

<br>

For the final project we will be working in a group of 3 team members whose details can be found below:

| Name   |      NetID      | 
|----------|:-------------:|
| Mohit Khanna | mkhanna2@illinois.edu |
| Eric A. Scuccimarra | eas7@illinois.edu |
| Shikhar Khanna | shikhar2@illinois.edu |

<br>

***

# Project Title

<br>

Understanding the effects of Seasonal and Environmental variables to predict the short term demand of bikes.

<br>

***

# Description of the dataset

<br>

Bike-sharing rental process is highly correlated to the environmental and seasonal settings. For instance, weather conditions, precipitation, day of week, season, hour of the day, etc. can affect the rental behaviors. The core data set is related to the two-year historical log corresponding to years 2011 and 2012 from Capital Bikeshare system, Washington D.C., USA which is publicly available in http://capitalbikeshare.com/system-data. The data is aggregated on daily basis and then extracted and added the corresponding weather and seasonal information.

The dataset consist of following variables:

	- instant: record index
	- dteday : date
	- season : season (1:springer, 2:summer, 3:fall, 4:winter)
	- yr : year (0: 2011, 1:2012)
	- mnth : month ( 1 to 12)
	- holiday : weather day is holiday or not (extracted from http://dchr.dc.gov/page/holiday-schedule)
	- weekday : day of the week
	- workingday : if day is neither weekend nor holiday is 1, otherwise is 0.
	+ weathersit : 
		- 1: Clear, Few clouds, Partly cloudy, Partly cloudy
		- 2: Mist + Cloudy, Mist + Broken clouds, Mist + Few clouds, Mist
		- 3: Light Snow, Light Rain + Thunderstorm + Scattered clouds, Light Rain + Scattered clouds
		- 4: Heavy Rain + Ice Pallets + Thunderstorm + Mist, Snow + Fog
	- temp : Normalized temperature in Celsius. The values are divided to 41 (max)
	- atemp: Normalized feeling temperature in Celsius. The values are divided to 50 (max)
	- hum: Normalized humidity. The values are divided to 100 (max)
	- windspeed: Normalized wind speed. The values are divided to 67 (max)
	- casual: count of casual users
	- registered: count of registered users
	- cnt: count of total rental bikes including both casual and registered

<br>

We will use this dataset to predict the count of rental bikes(cnt variable) based upon the features we have seen above.


<br>

***

# Background

<br>

Bike sharing systems are new generation of traditional bike rentals where whole process from membership, rental and return back has become automatic. Through these systems, user is able to easily rent a bike from a particular position and return back at another position. Currently, there are about over 500 bike-sharing programs around the world which is composed of over 500 thousands bicycles. Today, there exists great interest in these systems due to their important role in traffic, environmental and health issues. 


***Source:*** The core data set is related to the two-year historical log corresponding to years 2011 and 2012 from Capital Bikeshare system, Washington D.C., USA which is publicly available in http://capitalbikeshare.com/system-data.

<br>

***

# Motivation 

<br>

To understand how variables in the bike sharing dataset combine in complex relationships to predict the demand for bike and how attributes like like weather and time affect a quantitative outcome of bike rentals. The quantification of the relationship will assist in forecasting demand for bike rentals in the short term.

<br>

***

# Data Import

<br>

```{r}
bike_sharing_data<-read.csv(file="day.csv",header=TRUE, sep=",") # reading the dataset into memory
```

<br>

The dataset has following dimensions:

```{r}
dim(bike_sharing_data) # dimensions of the dataset
```

<br>

First few observation of target variable:

```{r}
head(bike_sharing_data$cnt) # printing first few rows of our target variable which is count of bikes for every day
```


***

