---
title: "Week 10- Homework Solutions"
author: "Mohit Khanna"
date: "13/07/2019"
output:
  html_document: 
    toc: yes
  pdf_document: default
urlcolor: cyan
---

***

```{r setup, echo = FALSE, message = FALSE, warning = FALSE}
options(scipen = 1, digits = 4, width = 80, fig.align = "center")
```

## Exercise 1 (Simulating Wald and Likelihood Ratio Tests)

In this exercise we will investigate the distributions of hypothesis tests for logistic regression. For this exercise, we will use the following predictors.

```{r}
sample_size = 150
set.seed(420)
x1 = rnorm(n = sample_size)
x2 = rnorm(n = sample_size)
x3 = rnorm(n = sample_size)
```

Recall that

$$
p({\bf x}) = P[Y = 1 \mid {\bf X} = {\bf x}]
$$

Consider the true model

$$
\log\left(\frac{p({\bf x})}{1 - p({\bf x})}\right) = \beta_0 + \beta_1 x_1
$$

where

- $\beta_0 = 0.4$
- $\beta_1 = -0.35$

**(a)** To investigate the distributions, simulate from this model 2500 times. To do so, calculate 

$$
P[Y = 1 \mid {\bf X} = {\bf x}]
$$ 

for an observation, and then make a random draw from a Bernoulli distribution with that success probability. (Note that a Bernoulli distribution is a Binomial distribution with parameter $n = 1$. There is no direction function in `R` for a Bernoulli distribution.)

Each time, fit the model:

$$
\log\left(\frac{p({\bf x})}{1 - p({\bf x})}\right) = \beta_0 + \beta_1 x_1 + \beta_2 x_2 + \beta_3 x_3
$$

Store the test statistics for two tests:

- The Wald test for $H_0: \beta_2 = 0$, which we say follows a standard normal distribution for "large" samples
- The likelihood ratio test for $H_0: \beta_2 = \beta_3 = 0$, which we say follows a $\chi^2$ distribution (with some degrees of freedom) for "large" samples

<br>

***Solution:***

```{r}
sim_logistic_data = function(sample_size = sample_size, beta_0, beta_1) {
  eta = beta_0 + beta_1 * x1
  p = 1 / (1 + exp(-eta))
  y = rbinom(n = sample_size, size = 1, prob = p)
  data.frame(y, x1,x2,x3)
}
test_stat_beta2=rep(0,2500)
likelihood_test_stat=rep(0,2500)

for (i in 1:2500){
  dataset=sim_logistic_data(sample_size = sample_size, beta_0 = 0.4, beta_1 = -0.35)
  sim_mod= glm(y~., data=dataset, family = "binomial")
  test_stat_beta2[i]=summary(sim_mod)[["coefficients"]][3, 3]
  true_mod=glm(y~x1,data=dataset, family = "binomial" )
  anova_test=anova(true_mod, sim_mod, test="LRT")
  likelihood_test_stat[i]=anova_test[["Deviance"]][2]
}
```

<br>

**(b)** Plot a histogram of the empirical values for the Wald test statistic. Overlay the density of the true distribution assuming a large sample.

<br>

***Solution:***

```{r}
hist(test_stat_beta2, prob=TRUE,xlab = "Test statistic for beta_2", main = "Histogram of empirical values for the Wald test statistic", border = "black", col="dodgerblue")
curve(dnorm(x, mean=0, sd=1), 
      col="orange", lwd=2, add=TRUE)
```

<br>

**(c)** Use the empirical results for the Wald test statistic to estimate the probability of observing a test statistic larger than 1. Also report this probability using the true distribution of the test statistic assuming a large sample.

<br>

***Solution:***

Probability of observing a test statistic larger than 1 using the empirical results for the Wald test statistic

```{r}
mean(test_stat_beta2>1)
```

<br>

Probability of observing a test statistic larger than 1 using the true distribution of the test statistic

```{r}
pnorm(1, mean=0, sd=1, lower.tail = FALSE)
```

<br>

**(d)** Plot a histogram of the empirical values for the likelihood ratio test statistic. Overlay the density of the true distribution assuming a large sample.

<br>

***Solution:***

```{r}
hist(likelihood_test_stat, prob=TRUE,xlab = "Test statistic for Likelihood ratio test", main = "Histogram of empirical values for the likelihood ratio test statistic", border = "black", col="dodgerblue")

curve( dchisq(x, df=2), col='orange', add=TRUE, lwd=3)

```

<br>

**(e)** Use the empirical results for the likelihood ratio test statistic to estimate the probability of observing a test statistic larger than 5. Also report this probability using the true distribution of the test statistic assuming a large sample.

<br>

***Solution:***

Probability of observing a test statistic larger than 5 using the empirical results for the likelihood ratio test statistic

```{r}
mean(likelihood_test_stat>5)
```

<br>

Probability of observing a test statistic larger than 5 using the true distribution of the likelihood ratio test statistic 

```{r}
pchisq(5, df=2, lower.tail = FALSE)
```


<br>

**(f)** Repeat **(a)**-**(e)** but with simulation using a smaller sample size of 10. Based on these results, is this sample size large enough to use the standard normal and $\chi^2$ distributions in this situation? Explain.

```{r}
sample_size = 10
set.seed(420)
x1 = rnorm(n = sample_size)
x2 = rnorm(n = sample_size)
x3 = rnorm(n = sample_size)
```

<br>

***Solution:***

```{r, warning=FALSE}
sim_logistic_data = function(sample_size = sample_size, beta_0, beta_1) {
  eta = beta_0 + beta_1 * x1
  p = 1 / (1 + exp(-eta))
  y = rbinom(n = sample_size, size = 1, prob = p)
  data.frame(y, x1,x2,x3)
}
test_stat_beta2=rep(0,2500)
likelihood_test_stat=rep(0,2500)

for (i in 1:2500){
  dataset=sim_logistic_data(sample_size = sample_size, beta_0 = 0.4, beta_1 = -0.35)
  sim_mod= glm(y~., data=dataset, family = "binomial")
  test_stat_beta2[i]=summary(sim_mod)[["coefficients"]][3, 3]
  true_mod=glm(y~x1,data=dataset, family = "binomial" )
  anova_test=anova(true_mod, sim_mod, test="LRT")
  likelihood_test_stat[i]=anova_test[["Deviance"]][2]
}
```

<br>

```{r}
hist(test_stat_beta2, prob=TRUE,xlab = "Test statistic for beta_2", main = "Histogram of empirical values for the Wald test statistic", border = "black", col="dodgerblue")
curve(dnorm(x, mean=0, sd=1), 
      col="orange", lwd=2, add=TRUE)
```

<br>

Probability of observing a test statistic larger than 1 using the empirical results for the Wald test statistic

```{r}
mean(test_stat_beta2>1)
```

<br>

Probability of observing a test statistic larger than 1 using the true distribution of the test statistic

```{r}
pnorm(1, mean=0, sd=1, lower.tail = FALSE)
```

<br>


```{r}
hist(likelihood_test_stat, prob=TRUE,xlab = "Test statistic for Likelihood ratio test", main = "Histogram of empirical values for the likelihood ratio test statistic", border = "black", col="dodgerblue")

curve( dchisq(x, df=2), col='orange', add=TRUE, lwd=3)

```

<br>

Probability of observing a test statistic larger than 5 using the empirical results for the likelihood ratio test statistic

```{r}
mean(likelihood_test_stat>5)
```

<br>

Probability of observing a test statistic larger than 5 using the true distribution of the likelihood ratio test statistic 

```{r}
pchisq(5, df=2, lower.tail = FALSE)
```
<br>

From the above plots and results it is clearly visible that the sample size of 10 is not large enough since the empirical results from Wald test and likelihood ratio test are not even close to the true distribution of those metrics. When the sample size was 150 we got results for these metrics which were really close to the true distribution thus this clearly proves that decreasing the sample sizes produces results which does not comply with true distribution.

<br>

***

## Exercise 2 (Surviving the Titanic)

For this exercise use the `ptitanic` data from the `rpart.plot` package. (The `rpart.plot` package depends on the `rpart` package.) Use `?rpart.plot::ptitanic` to learn about this dataset. We will use logistic regression to help predict which passengers aboard the [Titanic](https://en.wikipedia.org/wiki/RMS_Titanic) will survive based on various attributes.

```{r, message = FALSE, warning = FALSE}
# install.packages("rpart")
# install.packages("rpart.plot")
library(rpart)
library(rpart.plot)
data("ptitanic")
```

For simplicity, we will remove any observations with missing data. Additionally, we will create a test and train dataset.

```{r}
ptitanic = na.omit(ptitanic)
set.seed(42)
trn_idx = sample(nrow(ptitanic), 300)
ptitanic_trn = ptitanic[trn_idx, ]
ptitanic_tst = ptitanic[-trn_idx, ]
```

**(a)** Consider the model

$$
\log\left(\frac{p({\bf x})}{1 - p({\bf x})}\right) = \beta_0 + \beta_1 x_1 + \beta_2 x_2 + \beta_3 x_3 + \beta_4 x_4 + \beta_5 x_3x_4
$$

where

$$
p({\bf x}) = P[Y = 1 \mid {\bf X} = {\bf x}]
$$

is the probability that a certain passenger survives given their attributes and

- $x_1$ is a dummy variable that takes the value $1$ if a passenger was 2nd class.
- $x_2$ is a dummy variable that takes the value $1$ if a passenger was 3rd class.
- $x_3$ is a dummy variable that takes the value $1$ if a passenger was male.
- $x_4$ is the age in years of a passenger.

Fit this model to the training data and report its deviance.

<br>

***Solution:***

```{r}
tita_mod=glm(survived~pclass+sex+age+age:sex, data= ptitanic_trn, family="binomial")
deviance(tita_mod)
```

<br>

**(b)** Use the model fit in **(a)** and an appropriate statistical test to determine if class played a significant role in surviving on the Titanic. Use $\alpha = 0.01$. Report:

- The null hypothesis of the test
- The test statistic of the test
- The p-value of the test
- A statistical decision
- A practical conclusion

<br>

***Solution:***

***The null hypothesis of the test:***

$$\beta_1 = \beta_2=0$$

<br>

***The test statistic of the test:***

```{r}
tita_mod_null=glm(survived~sex+age+age:sex, data= ptitanic_trn, family="binomial")
anova(tita_mod_null, tita_mod, test="LRT")[["Deviance"]][2]
```

<br>

***The p-value of the test:***

```{r}
anova(tita_mod_null, tita_mod, test="LRT")[["Pr(>Chi)"]][2]
```

<br>

***A statistical decision:***

At alpha level of 0.01 we are going to reject the null hypothesis

<br>

***A practical conclusion:***

Class is a significant variable in determining who survived on Titanic.

<br>

**(c)** Use the model fit in **(a)** and an appropriate statistical test to determine if an interaction between age and sex played a significant role in surviving on the Titanic. Use $\alpha = 0.01$. Report:

- The null hypothesis of the test
- The test statistic of the test
- The p-value of the test
- A statistical decision
- A practical conclusion

<br>

***Solution:***

<br>

***The null hypothesis of the test***

The interaction between age and sex are insignificant and does not add any predictive power to the model

<br>

***The test statistic of the test:***

```{r}
tita_mod_wint=glm(survived~pclass+sex+age, data= ptitanic_trn, family="binomial")
anova(tita_mod_wint,tita_mod, test="LRT")[["Deviance"]][2]
```

<br>

***The p-value of the test:***

```{r}
anova(tita_mod_wint,tita_mod, test="LRT")[["Pr(>Chi)"]][2]
```

<br>

***A statistical decision:***

At alpha level of 0.01 we are going to reject the null hypothesis

<br>

***A practical conclusion:***

The interaction of age and sex is a significant term in the model.

<br>

**(d)** Use the model fit in **(a)** as a classifier that seeks to minimize the misclassifications rate. Classify each of the passengers in the test data set. Report the misclassification rate, the sensitivity, and the specificity of this classifier. (Use survived as the positive class.)

<br>

***Solution:***

The miss classification rate is :

```{r}
predict_test= predict(tita_mod, newdata = ptitanic_tst, type="response")
predict_test_class=ifelse(predict_test>0.5, "survived", "died")
1-mean(predict_test_class==ptitanic_tst$survived)
```

<br>

The sensitivity is :

```{r}
make_conf_mat = function(predicted, actual) {
  table(predicted = predicted, actual = actual)
}

# function to get sensitivity
get_sens = function(conf_mat) {
  conf_mat[2, 2] / sum(conf_mat[, 2])
}

# function to get specificity
get_spec =  function(conf_mat) {
  conf_mat[1, 1] / sum(conf_mat[, 1])
}

conf_mat_50 = make_conf_mat(predicted = predict_test_class, actual = ptitanic_tst$survived)

```

<br>

```{r}
get_sens(conf_mat_50)
```

<br>

The specificity is:
```{r}
get_spec(conf_mat_50)
```

<br>

***

## Exercise 3 (Breast Cancer Detection)

For this exercise we will use data found in [`wisc-train.csv`](wisc-train.csv) and [`wisc-test.csv`](wisc-test.csv), which contain train and test data, respectively. `wisc.csv` is provided but not used. This is a modification of the Breast Cancer Wisconsin (Diagnostic) data set from the UCI Machine Learning Repository. Only the first 10 feature variables have been provided. (And these are all you should use.)

- [UCI Page](https://archive.ics.uci.edu/ml/datasets/Breast+Cancer+Wisconsin+(Diagnostic))
- [Data Detail](https://archive.ics.uci.edu/ml/machine-learning-databases/breast-cancer-wisconsin/wdbc.names)

You should consider coercing the response to be a factor variable if it is not stored as one after importing the data.

**(a)** The response variable `class` has two levels: `M` if a tumor is malignant, and `B` if a tumor is benign. Fit three models to the training data.

- An additive model that uses `radius`, `smoothness`, and `texture` as predictors
- An additive model that uses all available predictors
- A model chosen via backwards selection using AIC. Use a model that considers all available predictors as well as their two-way interactions for the start of the search.

For each, obtain a 5-fold cross-validated mis classification rate using the model as a classifier that seeks to minimize the mis classification rate. Based on this, which model is best? Relative to the best, are the other two under-fitting or over fitting? Report the test mis classification rate for the model you picked as the best.

<br>

***Solution:***

```{r}
wisc_train <- read.csv(file="wisc-train.csv",header=TRUE, sep=",")
wisc_test <- read.csv(file="wisc-test.csv",header=TRUE, sep=",")
is.factor(wisc_train$class)
```


```{r, warning=FALSE}
mod1=glm(class~radius+smoothness+texture, data= wisc_train, family="binomial")
mod2=glm(class~., data= wisc_train, family="binomial")

mod3_start=glm(class~.^2, data= wisc_train, family="binomial")
mod3=step(mod3_start, trace=0, direction="backward")


boot::cv.glm(wisc_train, mod1, K = 5)$delta[1]
boot::cv.glm(wisc_train, mod2, K = 5)$delta[1]
boot::cv.glm(wisc_train, mod3, K = 5)$delta[1]
```

<br>

From the above results we can see that model 1 has the lowest mis classification rate and model 2 and model 3 are over-fitting relative to model 1

<br>

The miss classification rate is :

```{r}
predict_test= predict(mod1, newdata = wisc_test, type="response")
predict_test_class=ifelse(predict_test>0.5, "M", "B")
1-mean(predict_test_class==wisc_test$class)
```


<br>

**(b)** In this situation, simply minimizing mis classification might be a bad goal since false positives and false negatives carry very different consequences. Consider the `M` class as the "positive" label. Consider each of the probabilities stored in `cutoffs` in the creation of a classifier using the **additive** model fit in **(a)**.


That is, consider each of the values stored in `cutoffs` as $c$. Obtain the sensitivity and specificity in the test set for each of these classifiers. Using a single graphic, plot both sensitivity and specificity as a function of the cutoff used to create the classifier. Based on this plot, which cutoff would you use? (0 and 1 have not been considered for coding simplicity. If you like, you can instead consider these two values.)

$$
\hat{C}(\bf x) = 
\begin{cases} 
      1 & \hat{p}({\bf x}) > c \\
      0 & \hat{p}({\bf x}) \leq c 
\end{cases}
$$

<br>

***Solution:***

```{r}
cutoffs = seq(0.01, 0.99, by = 0.01)
sensitivity_val=rep(0, length(cutoffs))
specificity_val=rep(0, length(cutoffs))
i=1
predict_test= predict(mod1, newdata = wisc_test, type="response")
for (cut in cutoffs){
  predict_test_class=ifelse(predict_test>cut, "M", "B")
  conf_mat_50 = make_conf_mat(predicted = predict_test_class, actual = wisc_test$class)
  sensitivity_val[i]=get_sens(conf_mat_50)
  specificity_val[i]=get_spec(conf_mat_50)
  i=i+1
}
```

<br>

```{r}
plot(cutoffs, sensitivity_val, col="darkorange",
       xlab="Different cut off values",
      ylab="Sensitivity and Specificity",
       main="Effect of cutoff values on sensitivity and specificity",
       pch=20,
       cex=5,type="l", ylim=c(0.3, 1.0))

lines(cutoffs, specificity_val, col="dodgerblue", cex=5, pch=20)
legend("bottomright", c("Sensitivity", "Specificity"), lty = c(1, 1), lwd = 2,
         col = c("darkorange", "dodgerblue")) 
```


***Based on the above plot a cutoff value around 0.7 should be a good one.***


<br>

***
